---
title: "Reel Intelligence: MovieLens Clustering"
subtitle: "Final Project Report, Spring 2026"
author:
  - Ali Jabir
  - Isis Martínez
  - Susan Peck
date: "May 2, 2026"
format:
  pdf:
    pdf-engine: pdflatex
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{float}
  - \usepackage{setspace}
  - \usepackage{graphicx}
  - \usepackage{caption}
  - \usepackage{hyperref}
bibliography: refs.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5.5,
  dpi = 120
)

library(dplyr)
library(ggplot2)
library(knitr)
library(scales)


source("compare_results_helpers.R")

compare_results <- build_compare_results_data(
  cache_root = file.path("..", "output", "benchmark_cache")
)

plot_df <- compare_results$plot_df
best_models_df <- compare_results$best_models_df
best_table <- compare_results$best_table
overall_best <- compare_results$overall_best
ari_df <- compare_results$ari_df
ari_summary <- compare_results$ari_summary
```


\begin{center}
\textbf{Clustering Analysis (Math 252)}\\
Professor: Dr. Cristina Tortora\\
Department of Statistics\\
San Jose State University
\end{center}

# Abstract

This study evaluates clustering structure in user embeddings learned from the MovieLens 1M dataset. We generate embeddings using a multilayer perceptron and examine three embedding dimensions (64, 128, and 256). Four clustering methods are compared: k-means, probabilistic distance (PD) clustering, reduced k-means, and factorial probabilistic distance clustering (FPDC). Because no ground-truth labels are available, evaluation is based on internal validation metrics, bootstrap stability, partition agreement, and runtime.

The results provide only modest evidence of clustering structure in the learned embeddings. Reduced K-means produces the strongest and most consistent hard partitions, particularly in the 128-dimensional embedding space. K-means recovers similar structure but is sensitive to the choice of tuning criterion. In contrast, PD clustering and FPDC do not yield stable or well-separated partitions under the evaluation framework used in this study. Across methods, the 128-dimensional embeddings provide the clearest clustering signal.

Overall, the findings suggest that while embedding-based representations of user behavior contain some clustering tendency, the resulting structure is weak and sensitive to modeling choices, limiting strong interpretation of distinct user groups.

```{r child="sections/01-introduction.Rmd"}
```

```{r child="sections/02-background.Rmd"}
```

```{r child="sections/03-dataset.Rmd"}
```

```{r child="sections/04-methods.Rmd"}
```

```{r child="sections/05-evaluation.Rmd"}
```

```{r child="sections/06-results.Rmd"}
```

```{r child="sections/07-conclusion.Rmd"}
```

```{r child="sections/08-appendix.Rmd"}
```

# References
