---
title: "MovieLens 1M: Clustering the Embeddings"
author: ""
date: "`r format(Sys.Date())`"
output:
  pdf_document:
    toc: true
    number_sections: true
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 6,
  dpi = 120
)
set.seed(42)
```

# Overview


# 1. Requirements
```{r libraries}
required_packages <- c(
  "readr", "dplyr", "tidyr", "ggplot2",
  "cluster", "clusterCrit", "factoextra"
)

installed <- rownames(installed.packages())
missing_pkgs <- setdiff(required_packages, installed)
if (length(missing_pkgs) > 0) {
  install.packages(missing_pkgs)
}

library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(cluster)
library(clusterCrit)
library(factoextra)
```

# 2. Load embeddings files
```{r download-and-read}
embedding_paths <- c(
  "output/embeddings/user_embeddings_64.csv"  , "output/embeddings/user_embeddings_128.csv",
   "output/embeddings/user_embeddings_256.csv"
)

k_vals <- 2:20
all_results <- list()

dir.create("output/kmeans_plots", recursive = TRUE, showWarnings = FALSE)
dir.create("output/kmeans_cache", recursive = TRUE, showWarnings = FALSE)
dir.create("output/kmeans_results", recursive = TRUE, showWarnings = FALSE)
```



# 3. Choose Number of Clusters
```{r load-saved}
load_kmeans_cached_results <- function(embedding_path) {
  embeddings <- read_csv(embedding_path, show_col_types = FALSE)
  
  user_ids <- embeddings[[1]]
  embedding_matrix <- embeddings[, -1]
  embedding_matrix <- scale(as.matrix(embedding_matrix))
  
  embedding_tag <- tools::file_path_sans_ext(basename(embedding_path))
  full_file <- file.path(
    "output/kmeans_cache", paste0("kmeans_full_result_", 
                                  embedding_tag, ".rds")
    )

if (file.exists(full_file)) {
  all_results[[embedding_tag]] <- readRDS(full_file)
  next
}
  
  dmat <- dist(embedding_matrix)
  
  kmeans_file <- file.path("output/kmeans_cache", paste0("kmeans_result_", embedding_tag, ".rds"))
  all_file    <- file.path("output/kmeans_cache", paste0("k_results_all_", embedding_tag, ".rds"))
  mq_file     <- file.path("output/kmeans_cache", paste0("k_results_mq_", embedding_tag, ".rds"))
  
  if (!file.exists(kmeans_file) || !file.exists(all_file) || !file.exists(mq_file)) {
    return(NULL)
  }
  
  kmeans_result <- readRDS(kmeans_file)
  k_results_all <- readRDS(all_file)
  k_results_mq  <- readRDS(mq_file)
  
  best_k_sil <- k_results_mq$k[which.max(k_results_mq$Silhouette)]
  best_k_ch  <- k_results_mq$k[which.max(k_results_mq$CH)]
  
  elbow_info <- find_elbow_k(k_results_mq$k, k_results_mq$WSS)
  best_k_elbow <- elbow_info$elbow_k
  final_k <- kmeans_result$centers |> nrow()
  
  cluster_df <- data.frame(
    UserID = user_ids,
    Cluster = kmeans_result$cluster
  )
  
  final_sil <- silhouette(kmeans_result$cluster, dmat)
  final_avg_sil <- mean(final_sil[, 3])
  
  pca <- prcomp(embedding_matrix)
  var_explained <- summary(pca)$importance[2, 1:2]
  
  pca_df <- data.frame(
    PC1 = pca$x[, 1],
    PC2 = pca$x[, 2],
    Cluster = as.factor(kmeans_result$cluster)
  )
  
  list(
    embedding_path = embedding_path,
    embedding_tag = embedding_tag,
    embedding_matrix = embedding_matrix,
    user_ids = user_ids,
    dmat = dmat,
    k_results_mq = k_results_mq,
    k_results_all = k_results_all,
    best_k_sil = best_k_sil,
    best_k_ch = best_k_ch,
    best_k_elbow = best_k_elbow,
    elbow_distances = elbow_info$distances,
    final_k = final_k,
    kmeans_result = kmeans_result,
    cluster_df = cluster_df,
    final_sil = final_sil,
    final_avg_sil = final_avg_sil,
    pca_df = pca_df,
    var_explained = var_explained
  )
}
```


```{r eval-funct}
# KMEANS: CHOOSE NUMBER OF CLUSTERS
#* We evaluate k = 2 to 20 using:
  #* WSS (elbow method)
  #* average silhouette width
  #* Calinski-Harabasz (CH) index
# Function to evaluate one k-means algorithm across all k
# KMEANS: CHOOSE NUMBER OF CLUSTERS
# We evaluate k = 2 to 20 using:
# - WSS (elbow method)
# - average silhouette width
# - Calinski-Harabasz (CH) index



evaluate_kmeans <- function(x, k_vals, dmat, algorithm = "Hartigan-Wong",
                            nstart = 10, iter.max = 100) {
  
  wss <- numeric(length(k_vals))
  sil_scores <- numeric(length(k_vals))
  ch_scores <- numeric(length(k_vals))
  
  for (i in seq_along(k_vals)) {
    k <- k_vals[i]
    
    km <- kmeans(
      x,
      centers = k,
      nstart = nstart,
      iter.max = iter.max,
      algorithm = algorithm
    )
    
    wss[i] <- km$tot.withinss
    
    sil <- silhouette(km$cluster, dmat)
    sil_scores[i] <- mean(sil[, 3])
    
    ch_scores[i] <- intCriteria(
      traj = x,
      part = as.integer(km$cluster),
      crit = "Calinski_Harabasz"
    )$calinski_harabasz
  }
  
  data.frame(
    Algorithm = algorithm,
    k = k_vals,
    WSS = wss,
    Silhouette = sil_scores,
    CH = ch_scores
  )
}

# Detect elbow from WSS curve using maximum distance to the line
find_elbow_k <- function(k, wss) {
  pts <- cbind(k, wss)
  p1 <- pts[1, ]
  p2 <- pts[nrow(pts), ]
  
  point_line_distance <- function(p, a, b) {
    num <- abs((b[2] - a[2]) * p[1] - (b[1] - a[1]) * p[2] + b[1] * a[2] - b[2] * a[1])
    den <- sqrt((b[2] - a[2])^2 + (b[1] - a[1])^2)
    num / den
  }
  
  distances <- apply(pts, 1, point_line_distance, a = p1, b = p2)
  elbow_index <- which.max(distances)
  
  list(
    elbow_k = k[elbow_index],
    distances = distances
  )
}


```
#. SELECT FINAL K
Choose final k from MacQueen results -- Show all algorithms agree --> Emphasize robustness, not superiority

* choose k using the elbow method on WSS as the primary criterion
* use silhouette and CH as supporting diagnostics
* check cluster size balance before finalizing
* set k manually after inspecting results if needed

# 4. Loop over all embeddings
```{r loop}

# Run all three algorithms
for (embedding_path in embedding_paths) {
  
  if (!file.exists(embedding_path)) {
    cat("Missing file:", embedding_path, "\n")
    next
  }
  
  cat("\nProcessing:", embedding_path, "\n")
  
  embeddings <- read_csv(embedding_path, show_col_types = FALSE)
  user_ids <- embeddings[[1]]
  embedding_matrix <- embeddings[, -1]
  embedding_matrix <- scale(as.matrix(embedding_matrix))
  
  embedding_tag <- tools::file_path_sans_ext(basename(embedding_path))
  
  kmeans_file <- file.path("output/kmeans_cache", paste0("kmeans_result_", embedding_tag, ".rds"))
  all_file    <- file.path("output/kmeans_cache", paste0("k_results_all_", embedding_tag, ".rds"))
  mq_file     <- file.path("output/kmeans_cache", paste0("k_results_mq_", embedding_tag, ".rds"))
  
  if (file.exists(kmeans_file) && file.exists(all_file) && file.exists(mq_file)) {
    cat("Loading cached results for:", embedding_tag, "\n")
    
    cached_res <- load_kmeans_cached_results(embedding_path)
    all_results[[embedding_tag]] <- cached_res
    next
  }
  
  cat("Running k-means for:", embedding_tag, "\n")
  
  dmat <- dist(embedding_matrix)
  
  k_results_hw <- evaluate_kmeans(embedding_matrix, k_vals, dmat, "Hartigan-Wong")
  k_results_lloyd <- evaluate_kmeans(embedding_matrix, k_vals, dmat, "Lloyd")
  k_results_mq <- evaluate_kmeans(embedding_matrix, k_vals, dmat, "MacQueen")
  
  k_results_all <- bind_rows(k_results_hw, k_results_lloyd, k_results_mq)
  
  best_k_sil <- k_results_mq$k[which.max(k_results_mq$Silhouette)]
  best_k_ch  <- k_results_mq$k[which.max(k_results_mq$CH)]
  
  elbow_info <- find_elbow_k(k_results_mq$k, k_results_mq$WSS)
  best_k_elbow <- elbow_info$elbow_k
  final_k <- best_k_elbow
  
  set.seed(123)
  kmeans_result <- kmeans(
    embedding_matrix,
    centers = final_k,
    nstart = 100,
    iter.max = 100,
    algorithm = "MacQueen"
  )
  
  cluster_df <- data.frame(
    UserID = user_ids,
    Cluster = kmeans_result$cluster
  )
  
  final_sil <- silhouette(kmeans_result$cluster, dmat)
  final_avg_sil <- mean(final_sil[, 3])
  
  pca <- prcomp(embedding_matrix)
  var_explained <- summary(pca)$importance[2, 1:2]
  
  pca_df <- data.frame(
    PC1 = pca$x[, 1],
    PC2 = pca$x[, 2],
    Cluster = as.factor(kmeans_result$cluster)
  )
  
  all_results[[embedding_tag]] <- list(
    embedding_path = embedding_path,
    embedding_tag = embedding_tag,
    embedding_matrix = embedding_matrix,
    user_ids = user_ids,
    dmat = dmat,
    k_results_hw = k_results_hw,
    k_results_lloyd = k_results_lloyd,
    k_results_mq = k_results_mq,
    k_results_all = k_results_all,
    best_k_sil = best_k_sil,
    best_k_ch = best_k_ch,
    best_k_elbow = best_k_elbow,
    elbow_distances = elbow_info$distances,
    final_k = final_k,
    kmeans_result = kmeans_result,
    cluster_df = cluster_df,
    final_sil = final_sil,
    final_avg_sil = final_avg_sil,
    pca_df = pca_df,
    var_explained = var_explained
  )
  
  cat("Saved results for:", embedding_tag, "\n")
}

```
# 5. Show results for all embeddings
```{r results}
for (name in names(all_results)) {
  cat("\n==========", name, "==========\n")
  cat("Best k by elbow:", all_results[[name]]$best_k_elbow, "\n")
  cat("Best k by silhouette:", all_results[[name]]$best_k_sil, "\n")
  cat("Best k by CH:", all_results[[name]]$best_k_ch, "\n")
  cat("Final average silhouette:", all_results[[name]]$final_avg_sil, "\n")
  cat("Cluster sizes:\n")
  print(table(all_results[[name]]$cluster_df$Cluster))
}
```

# 6. Plot tuning results for all embeddings
```{r plots}
for (name in names(all_results)) {
  res <- all_results[[name]]
  
  p1 <- ggplot(res$k_results_all, aes(x = k, y = Silhouette, color = Algorithm)) +
    geom_line() +
    geom_point() +
    theme_minimal() +
    labs(title = paste("Silhouette by k -", name))
  
  print(p1)
  
  ggsave(
    file.path("output/kmeans_plots", paste0("silhouette_", name, ".png")),
    p1, width = 9, height = 6, dpi = 120
  )
  
  p2 <- ggplot(res$k_results_all, aes(x = k, y = CH, color = Algorithm)) +
    geom_line() +
    geom_point() +
    theme_minimal() +
    labs(title = paste("CH by k -", name))
  
  print(p2)
  
  ggsave(
    file.path("output/kmeans_plots", paste0("ch_", name, ".png")),
    p2, width = 9, height = 6, dpi = 120
  )
  
  p3 <- ggplot(res$k_results_all, aes(x = k, y = WSS, color = Algorithm)) +
    geom_line() + geom_point() + geom_vline(
      xintercept = res$best_k_elbow, linetype = "dashed", color = "black",
      linewidth = 1
      ) +
    annotate("text", x = res$best_k_elbow, 
             y = max(res$k_results_all$WSS, na.rm = TRUE),
             label = paste("Elbow k =", res$best_k_elbow),
             vjust = -0.5, color = "black") +
    theme_minimal() +  labs(title = paste("WSS by k (Elbow Method) -", name))
  
  print(p3)
  
  ggsave(
    file.path("output/kmeans_plots", paste0("wss_", name, ".png")),
    p3, width = 9, height = 6, dpi = 120
  )
}




```

# 7. Plot PCA for all embeddings
```{r pca}
for (name in names(all_results)) {
  res <- all_results[[name]]
  
  p_pca <- ggplot(res$pca_df, aes(x = PC1, y = PC2, color = Cluster)) +
    geom_point(alpha = 0.6) +
    theme_minimal() +
    labs(
      title = paste("K-Means Clusters (PCA Projection) -", name,
                    "k =", res$final_k),
      x = paste0("PC1 (", round(100 * res$var_explained[1], 1), "%)"),
      y = paste0("PC2 (", round(100 * res$var_explained[2], 1), "%)")
    )
  
  print(p_pca)
  
  ggsave(
    file.path("output/kmeans_plots", paste0("pca_", name, "_k", res$final_k, ".png")),
    p_pca, width = 9, height = 6, dpi = 120
  )
}
```

# 8.Choose best overall embeddings
```{r choose}
all_best <- lapply(all_results, function(res) {
  data.frame(
    embedding = res$embedding_tag,
    best_k = res$final_k,
    Elbow_k = res$best_k_elbow,
    Silhouette_at_elbow = res$final_avg_sil,
    CH_max = max(res$k_results_mq$CH, na.rm = TRUE)
  )
})

all_best_df <- bind_rows(all_best)
print(all_best_df)

best_overall_index <- with(all_best_df, order(-Silhouette_at_elbow, -CH_max))[1]
best_overall <- all_best_df[best_overall_index, ]
print(best_overall)

best_name <- best_overall$embedding
best_res <- all_results[[best_name]]

best_kmeans <- best_res$kmeans_result
best_cluster_df <- best_res$cluster_df
best_pca_df <- best_res$pca_df
best_final_k <- best_res$final_k
```

# 9. Save per-embedding results
```{r save}
#### 11. SAVE RESULTS #####

for (name in names(all_results)) {
  res <- all_results[[name]]
  
  write.csv(
    res$cluster_df,  file = file.path("output/kmeans_results", paste0("kmeans_clusters_", name, "_k", res$final_k, ".csv")), row.names = FALSE
    )

  write.csv(
    res$k_results_all, file = file.path("output/kmeans_results",  paste0("kmeans_k_selection_metrics_all_", name, ".csv")), row.names = FALSE
    )

  write.csv(
    res$k_results_mq,
    file = file.path("output/kmeans_results", paste0("kmeans_k_selection_metrics_macqueen_", name, ".csv")),
    row.names = FALSE
  )
  
  saveRDS(
    res$kmeans_result,
    file = file.path("output/kmeans_cache", paste0("kmeans_result_", name, ".rds"))
  )
  
  saveRDS(
    res$k_results_all,
    file = file.path("output/kmeans_cache", paste0("k_results_all_", name, ".rds"))
  )
  
  saveRDS(
    res$k_results_mq,
    file = file.path("output/kmeans_cache", paste0("k_results_mq_", name, ".rds"))
  )
  
  saveRDS(
  res$k_results_hw,
  file = file.path("output/kmeans_cache", paste0("k_results_hw_", name, ".rds"))
)

saveRDS(
  res$k_results_lloyd,
  file = file.path("output/kmeans_cache", paste0("k_results_lloyd_", name, ".rds"))
)
  
  saveRDS(
    res, file = file.path("output/kmeans_cache", paste0("kmeans_full_result_", name, ".rds"))
    )
  
}

```


