---
title: "MovieLens 1M: Clustering the Embeddings"
author: ""
date: "`r format(Sys.Date())`"
output:
  pdf_document:
    toc: true
    number_sections: true
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 6,
  dpi = 120
)
set.seed(42)
```

# Overview


# Requirements


```{r libraries}
required_packages <- c(
  "readr", "dplyr", "tidyr", "ggplot2",
  "cluster", "factoextra",
  "FPDclustering"
)

installed <- rownames(installed.packages())
missing_pkgs <- setdiff(required_packages, installed)
if (length(missing_pkgs) > 0) {
  install.packages(missing_pkgs)
}

library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(cluster)
library(factoextra)
library(clustertend)
library(FPDclustering)
```

# 1. Load embeddings files

```{r download-and-read}
embedding_paths <- c(
  "output/embeddings/user_embeddings_64.csv",
  "output/embeddings/user_embeddings_128.csv",
  "output/embeddings/user_embeddings_256.csv"
)

k_vals <- 2:20
all_results <- list()
```

# 2. Choose Number of Clusters

```{r choose-wss}



####################### 4. KMEANS: CHOOSE NUMBER OF CLUSTERS #######################
#* We evaluate k = 2 to 20 using:
  #* WSS (elbow method)
  #* average silhouette width
  #* Calinski-Harabasz (CH) index
# Function to evaluate one k-means algorithm across all k
evaluate_kmeans <- function(x, k_vals, dmat, algorithm = "Hartigan-Wong",
                            nstart = 10, iter.max = 100) {
  
  wss <- numeric(length(k_vals))
  sil_scores <- numeric(length(k_vals))
  ch_scores <- numeric(length(k_vals))
  
  for (i in seq_along(k_vals)) {
    k <- k_vals[i]
    
    km <- kmeans(
      x,
      centers = k,
      nstart = nstart,
      iter.max = iter.max,
      algorithm = algorithm
    )
    
    wss[i] <- km$tot.withinss
    
    sil <- silhouette(km$cluster, dmat)
    sil_scores[i] <- mean(sil[, 3])
    
    ch_scores[i] <- intCriteria(
      traj = as.matrix(x),
      part = as.integer(km$cluster),
      crit = "Calinski_Harabasz"
    )$calinski_harabasz
  }
  
  data.frame(
    Algorithm = algorithm,
    k = k_vals,
    WSS = wss,
    Silhouette = sil_scores,
    CH = ch_scores
  )
}

# Run all three algorithms
k_results_hw <- evaluate_kmeans(embedding_matrix, k_vals, dmat, "Hartigan-Wong")
k_results_lloyd <- evaluate_kmeans(embedding_matrix, k_vals, dmat, "Lloyd")
k_results_mq <- evaluate_kmeans(embedding_matrix, k_vals, dmat, "MacQueen")

# Combine results into one table
k_results_all <- dplyr::bind_rows(k_results_hw, k_results_lloyd, k_results_mq)

# Comparison Table for all results (wide to be easier to compare)
k_results_wide <- k_results_all %>% 
  pivot_wider(names_from = Algorithm, values_from = c(WSS, Silhouette, CH))

print(k_results_wide)
```



# Comparison plots

```{r kmeans}
# 3. PLOT K-SELECTION METRICS

ggplot(k_results_all, aes(x = k, y = Silhouette, color = Algorithm)) +
  geom_line() + geom_point() + theme_minimal() + 
  labs(title = "Silhouette by k and k-means algorithm")

ggplot(k_results_all, aes(x = k, y = CH, color = Algorithm)) +
  geom_line() + geom_point() + theme_minimal() +
  labs(title = "Calinski-Harabasz by k and k-means algorithm")

ggplot(k_results_all, aes(x = k, y = WSS, color = Algorithm)) +
  geom_line() + geom_point() + theme_minimal() +
  labs(title = "WSS by k and k-means algorithm")

# JUST MACQUEEN --> Use MacQueen for final results, Show all algorithms agree --> Emphasize robustness, not superiority
print(k_results_mq)


ggplot(k_results_mq, aes(x = k, y = Silhouette, color = Algorithm)) +
  geom_line() + geom_point() + theme_minimal() + 
  labs(title = "Silhouette by k and k-means algorithm")

ggplot(k_results_mq, aes(x = k, y = CH, color = Algorithm)) +
  geom_line() + geom_point() + theme_minimal() +
  labs(title = "Calinski-Harabasz by k and k-means algorithm")

ggplot(k_results_mq, aes(x = k, y = WSS, color = Algorithm)) +
  geom_line() + geom_point() + theme_minimal() +
  labs(title = "WSS by k and k-means algorithm")

```




# 4. SELECT FINAL K #######################
Choose final k from MacQueen results -- Show all algorithms agree --> Emphasize robustness, not superiority

* choose k using silhouette as primary criterion
* use CH as confirmation
* check cluster size balance before finalizing
* set k manually after inspecting results.

```{r mq}
# Choose final k from MacQueen results
best_k_sil <- k_results_mq$k[which.max(k_results_mq$Silhouette)]
best_k_ch  <- k_results_mq$k[which.max(k_results_mq$CH)]

cat("Best k by silhouette:", best_k_sil, "\n", "Best k by CH:", best_k_ch, "\n")

final_k <- best_k_sil

####################### 5. FIT FINAL K-MEANS MODEL #######################
set.seed(123)

kmeans_result <- kmeans(
  embedding_matrix,
  centers = final_k,
  nstart = 100, 
  iter.max = 100,
  algorithm = "MacQueen"
)

# Save cluster labels with user IDs
cluster_df <- data.frame(
  UserID = user_ids,
  Cluster = kmeans_result$cluster
)

# Check cluster sizes 
table(kmeans_result$cluster)

```

# 9. FINAL SILHOUETTE SCORE 
```{r}
final_sil <- silhouette(kmeans_result$cluster, dmat)
final_avg_sil <- mean(final_sil[, 3])

cat("Final average silhouette width:", final_avg_sil, "\n")

# Silhouette plot
plot(
  final_sil,
  main = paste("Silhouette Plot for k =", final_k),
  col = "blue",
  border = NA
)


```
# 4. Visualize with PCA


```{r pca-visual}
# For visualization, not for fitting k-means.
pca <- prcomp(embedding_matrix)

var_explained <- summary(pca)$importance[2, 1:2]

pca_df <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  Cluster = as.factor(kmeans_result$cluster)
)

ggplot(pca_df, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(alpha = 0.6) +
  theme_minimal() +
  labs(
    title = paste("K-Means Clusters (PCA Projection), k =", final_k),
    x = paste0("PC1 (", round(100 * var_explained[1], 1), "%)"),
    y = paste0("PC2 (", round(100 * var_explained[2], 1), "%)")
  )
```

```{r save}
####################### 11. SAVE RESULTS #######################

write.csv(cluster_df, "output/kmeans_cluster_assignments.csv", row.names = FALSE)
write.csv(k_results_all, "output/kmeans_k_selection_metrics_all.csv", row.names = FALSE)
write.csv(k_results_mq, "output/kmeans_k_selection_metrics_macqueen.csv", row.names = FALSE)

saveRDS(kmeans_result, "output/kmeans_result.rds")
saveRDS(k_results_all, "output/k_results_all.rds")
saveRDS(k_results_mq, "output/k_results_mq.rds")

```
