---
title: "MovieLens 1M: Reduced K-means Clustering the Embeddings"
author: ""
date: "`r format(Sys.Date())`"
output:
  pdf_document:
    toc: true
    number_sections: true
fontsize: 11pt
geometry: margin=1in
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 6,
  dpi = 120
)
set.seed(42)
```

```{r}
required_packages <- c(
  "readr", "dplyr", "ggplot2",
  "cluster", "factoextra", "clustrd"
)

installed <- rownames(installed.packages())
missing_pkgs <- setdiff(required_packages, installed)

if (length(missing_pkgs) > 0) {
  install.packages(missing_pkgs)
}

library(readr)
library(dplyr)
library(ggplot2)
library(cluster)
library(factoextra)
library(clustrd)

```

# 1. Load embeddings files

```{r}
embedding_path <- "output/embeddings/user_embeddings_128.csv"

embeddings <- read_csv(embedding_path)

user_ids <- embeddings[[1]]
embedding_matrix <- embeddings[, -1]

embedding_matrix <- as.matrix(embedding_matrix)

dim(embedding_matrix)
head(user_ids)
```

# Add caching

```{r cache_setup}
# Create folders for cached models/results and plots
dir.create("output/rkm_cache", recursive = TRUE, showWarnings = FALSE)
dir.create("output/rkm_plots", recursive = TRUE, showWarnings = FALSE)

# Helper to build a safe tag from embedding file name
embedding_tag <- tools::file_path_sans_ext(basename(embedding_path))
```

```{r}
k_range <- 2:10
q_range <- 2:6

embedding_paths <- c(
  "output/embeddings/user_embeddings_128.csv",
  "output/embeddings/user_embeddings_256.csv",
  "output/embeddings/user_embeddings_64.csv"
)

all_results <- list()

for (embedding_path in embedding_paths) {
  
  cat("\nProcessing:", embedding_path, "\n")
  
  embeddings <- read_csv(embedding_path)
  
  user_ids <- embeddings[[1]]
  embedding_matrix <- embeddings[, -1]
  embedding_matrix <- as.matrix(embedding_matrix)
  # embedding_matrix <- scale(embedding_matrix)
  
  embedding_tag <- tools::file_path_sans_ext(basename(embedding_path))
  
  dmat <- dist(embedding_matrix)
  
  results_grid <- expand.grid(k = k_range, q = q_range)
  results_grid$asw <- NA
  
  rkm_models <- vector("list", nrow(results_grid))
  
  for (i in seq_len(nrow(results_grid))) {
    k_val <- results_grid$k[i]
    q_val <- results_grid$q[i]
    
    cache_file <- file.path(
      "output/rkm_cache",
      paste0("rkm_", embedding_tag, "_k", k_val, "_q", q_val, ".rds")
    )
    
    if (file.exists(cache_file)) {
      cat("Loading cached result for", embedding_tag, "k =", k_val, "q =", q_val, "\n")
      cached <- readRDS(cache_file)
      model <- cached$model
      avg_sil <- cached$asw
    } else {
      cat("Running RKM for", embedding_tag, "k =", k_val, "q =", q_val, "\n")
      
      model <- cluspca(
        embedding_matrix,
        nclus = k_val,
        ndim = q_val,
        method = "RKM",
        rotation = "varimax"
      )
      
      sil <- silhouette(model$cluster, dmat)
      avg_sil <- mean(sil[, 3])
      
      saveRDS(
        list(
          model = model,
          asw = avg_sil,
          k = k_val,
          q = q_val,
          embedding_path = embedding_path
        ),
        cache_file
      )
    }
    
    results_grid$asw[i] <- avg_sil
    rkm_models[[i]] <- model

  best_index <- which.max(results_grid$asw)
  best_params <- results_grid[best_index, ]
  
  best_k <- best_params$k
  best_q <- best_params$q
  best_asw <- best_params$asw
  best_rkm <- rkm_models[[best_index]]
  
  cluster_df <- data.frame(
    UserID = user_ids,
    Cluster = best_rkm$cluster
  )
  
  all_results[[embedding_tag]] <- list(
    embedding_path = embedding_path,
    embedding_tag = embedding_tag,
    embedding_matrix = embedding_matrix,
    user_ids = user_ids,
    results_grid = results_grid,
    best_index = best_index,
    best_params = best_params,
    best_k = best_k,
    best_q = best_q,
    best_asw = best_asw,
    best_rkm = best_rkm,
    cluster_df = cluster_df,
    dmat = dmat
  )
  }
}
```

```{r}
# best_index <- which.max(results_grid$asw)
# best_params <- results_grid[best_index, ]

# best_k <- best_params$k
# best_q <- best_params$q
# best_asw <- best_params$asw

# best_rkm <- rkm_models[[best_index]]

# best_params
```

```{r}
for (name in names(all_results)) {
  res <- all_results[[name]]
  
  p_tuning <- ggplot(res$results_grid, aes(x = q, y = asw, color = factor(k), group = k)) +
    geom_line() +
    geom_point(size = 2) +
    theme_minimal() +
    labs(
      title = paste("Reduced K-Means Tuning Results -", name),
      x = "Reduced Dimension (q)",
      y = "Average Silhouette Width",
      color = "Clusters (k)"
    )
  
  print(p_tuning)
  
  ggsave(
    filename = file.path("output/rkm_plots", paste0("tuning_", name, ".png")),
    plot = p_tuning,
    width = 9,
    height = 6,
    dpi = 120
  )
}
```

```{r}
# cluster_df <- data.frame(
#   UserID = user_ids,
#   Cluster = best_rkm$cluster
# )

# head(cluster_df)
# table(cluster_df$Cluster)
```

```{r}
for (name in names(all_results)) {
  res <- all_results[[name]]
  
  best_sil <- silhouette(res$best_rkm$cluster, res$dmat)
  
  png(
    filename = file.path(
      "output/rkm_plots",
      paste0("silhouette_", name, "_k", res$best_k, "_q", res$best_q, ".png")
    ),
    width = 1200,
    height = 800,
    res = 120
  )
  
  plot(
    best_sil,
    main = paste("Reduced K-Means Silhouette Plot -", name,
                 "(k =", res$best_k, ", q =", res$best_q, ")"),
    col = "blue",
    border = NA
  )
  
  dev.off()
  
  # Also show inside report
  plot(
    best_sil,
    main = paste("Reduced K-Means Silhouette Plot -", name,
                 "(k =", res$best_k, ", q =", res$best_q, ")"),
    col = "blue",
    border = NA
  )
}
```

```{r}
for (name in names(all_results)) {
  res <- all_results[[name]]
  
  pca <- prcomp(res$embedding_matrix)
  
  pca_df <- data.frame(
    PC1 = pca$x[, 1],
    PC2 = pca$x[, 2],
    Cluster = as.factor(res$best_rkm$cluster)
  )
  
  p_pca <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Cluster)) +
    geom_point(alpha = 0.6) +
    theme_minimal() +
    labs(
      title = paste("Reduced K-Means Clusters (PCA Projection) -", name,
                    "k =", res$best_k, ", q =", res$best_q),
      x = "PC1",
      y = "PC2"
    )
  
  print(p_pca)
  
  ggsave(
    filename = file.path(
      "output/rkm_plots",
      paste0("pca_", name, "_k", res$best_k, "_q", res$best_q, ".png")
    ),
    plot = p_pca,
    width = 9,
    height = 6,
    dpi = 120
  )
}
```

```{r}
best_rkm$centroid
plot(best_rkm, cludesc = TRUE)
```

```{r}
```{r}
for (name in names(all_results)) {
  res <- all_results[[name]]
  
  write.csv(
    res$cluster_df,
    file = file.path(
      "output",
      paste0("rkm_clusters_", name, "_k", res$best_k, "_q", res$best_q, ".csv")
    ),
    row.names = FALSE
  )
  
  write.csv(
    res$results_grid,
    file = file.path(
      "output",
      paste0("rkm_tuning_results_", name, ".csv")
    ),
    row.names = FALSE
  )
  
  saveRDS(
    res$best_rkm,
    file = file.path(
      "output/rkm_cache",
      paste0("best_rkm_", name, "_k", res$best_k, "_q", res$best_q, ".rds")
    )
  )
}
```