---
title: "Reel Insight (working title)"
subtitle: "MovieLens 1M: Clustering the Embeddings"
author: "Ali Jabir, Isis Martinez, Susan Peck"
thanks: "Project repository available at: [https://github.com/syedalijabir/math252-project](https://github.com/syedalijabir/math252-project)."
date: "`r format(Sys.Date())`"
format:
  pdf:
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
  include: true
  fig-width: 4    # default width (in inches)
  fig-height: 3   # default height (in inches)
  abstract: "xxx"
  bibliography: references.bib
  body-footer: Math 252 - Cluster Analysis | Spring 2026 Project
---


## CODE: To be moved as needed
# Setup


```{r setup}
source("setup.R")
```


```{r kmeans}
# Load saved k-means results
kmeans_result <- readRDS("output/kmeans_result.rds")
k_results_all <- readRDS("output/k_results_all.rds")
k_results_mq <- readRDS("output/k_results_mq.rds")
```

```{r pd}
# Load saved PD results
pd_results <- readRDS("output/pd_results.rds")
pd_pca_df <- readRDS("output/pd_pca_df.rds")
```
# PD METHODS
The Within-Cluster Sum of Squares (WSS) was used for k-means clustering because it is directly tied to the algorithm’s objective function. However, WSS is not applicable to PD clustering, which does not optimize a centroid-based objective. Therefore, method-independent validation metrics such as silhouette score and Calinski–Harabasz index were used to evaluate and select the number of clusters for PD clustering.

```{r}
# Load saved interpretation outputs
base_path <- "output/cluster_interpretations"

rating_summary <- read_csv(file.path(base_path, "py_rating_summary.csv"), show_col_types = FALSE)
activity_summary <- read_csv(file.path(base_path, "py_activity_summary.csv"), show_col_types = FALSE)
top_genres <- read_csv(file.path(base_path, "py_top_genres.csv"), show_col_types = FALSE)
demographic_summary <- read_csv(file.path(base_path, "py_demographic_summary.csv"), show_col_types = FALSE)
cluster_profile <- read_csv(file.path(base_path, "py_cluster_profile.csv"), show_col_types = FALSE)
```

```{r}
# K-means model selection
print(k_results_mq)

# Cluster profile
print(cluster_profile)
```