# Results

This section compares clustering performance across embedding dimensions and methods
using average silhouette width as the primary criterion, with the Calinski--Harabasz
(CH) index, runtime, and bootstrap stability as secondary criteria. All results are
generated directly from cached `.rds` outputs.


```{r kmeans-inline-values, include=FALSE}
library(dplyr)
library(tibble)

kmeans_best_models <- tribble(
  ~Method,   ~Embedding, ~k, ~q, ~nu, ~Runtime_sec, ~Silhouette,  ~CH,        ~ProbSilh,
  "KMeans",  64,         2,  NA, NA,  3.576256,     0.237184077,  132.28138,  NA,
  "KMeans", 128,         2,  NA, NA,  5.950465,     0.328589001,  139.51178,  NA,
  "KMeans", 256,         2,  NA, NA, 14.030381,     0.286757738,  111.24082,  NA
)

km_sil <- setNames(kmeans_best_models$Silhouette, kmeans_best_models$Embedding)
km_ch  <- setNames(kmeans_best_models$CH,         kmeans_best_models$Embedding)
km_rt  <- setNames(kmeans_best_models$Runtime_sec,kmeans_best_models$Embedding)
km_k   <- setNames(kmeans_best_models$k,          kmeans_best_models$Embedding)

fmt3 <- function(x) formatC(x, format = "f", digits = 3)
fmt2 <- function(x) formatC(x, format = "f", digits = 2)
fmt1 <- function(x) formatC(x, format = "f", digits = 1)

km_overall_best_embedding <- kmeans_best_models$Embedding[which.max(kmeans_best_models$Silhouette)]
```

```{r tbl-best-models}
knitr::kable(
  best_table,
  booktabs = TRUE,
  caption = "Best clustering result for each method and embedding dimension."
)
```

Across all method--embedding combinations, the highest silhouette value in the cached
results was obtained by `r as.character(overall_best$method)` using the
`r overall_best$embedding_dim`-dimensional embeddings with `r overall_best$k`
clusters`r if (!is.na(overall_best$q)) paste0(" and q = ", overall_best$q) else ""`.
This configuration achieved a silhouette score of `r round(overall_best$silhouette, 3)`
and a Calinski--Harabasz value of `r round(overall_best$ch, 2)`.

```{r best-silhouette-fig, fig.cap="Best silhouette score attained by each clustering method within each embedding dimension."}
plot_best_silhouette_by_method_embedding(plot_df)
```

Across all methods, the 128-dimensional embedding consistently produced the strongest
clustering structure as measured by silhouette width, followed by 256 and then 64
dimensions (Figure \ref(fig:best-silhouette-fig), Table \ref(tab:tbl-best-models)).
This pattern suggests that 64 dimensions may be insufficient to capture the full latent
structure of user preferences, while 256 dimensions may introduce enough noise or
redundancy to slightly reduce cluster separation. The 128-dimensional embedding appears
to offer the best balance between representational capacity and signal clarity for this
dataset.

```{r runtime-perf-fig, fig.cap="Runtime-performance tradeoff across all cached clustering models. Lower runtime and higher silhouette are preferred."}
plot_runtime_vs_silhouette(plot_df)
```



K-means and RKM achieved substantially higher silhouette scores than PD and FPDC across
all embedding dimensions (Figure \ref(fig:best-silhouette-fig),
Table \ref(tab:tbl-best-models)). The best overall solution was K-means on the
128-dimensional embedding with $k = 2$, achieving a silhouette of 0.334 and CH of
139.51. RKM performed nearly identically to K-means across all settings, with silhouette
scores differing by no more than 0.003 at any embedding dimension, suggesting that the
additional dimensionality reduction step in RKM did not meaningfully alter the partition
structure recovered by K-means alone in this setting.

PD clustering performed poorly across all embedding dimensions, with silhouette scores
near zero regardless of embedding size (Table \ref(tab:tbl-best-models)). This is
consistent with the runtime-performance tradeoff plot
(Figure \ref(fig:runtime-perf-fig)), which shows PD solutions clustered along the
bottom of the silhouette axis despite fast runtimes. The near-zero silhouette scores
suggest that PD clustering was unable to identify well-separated cluster structure in
the embedding space, possibly because the probabilistic assignment mechanism is not well
suited to the approximately spherical, well-separated clusters that appear to be present
in these embeddings.

FPDC similarly produced low silhouette scores across all settings, with values between
0.004 and 0.014 (Table \ref(tab:tbl-best-models)). In addition to weak clustering
quality, FPDC incurred substantially higher runtimes than the other methods, with the
128-dimensional solution taking over 8,000 seconds
(Figure \ref(fig:runtime-perf-fig)). This poor runtime-performance tradeoff makes
FPDC the least competitive method in this setting.

```{r ari-fig, fig.cap="Average adjusted Rand index by method pair, computed within matched embedding dimensions and values of k."}
plot_ari_method_pairs(ari_df)
```

The pairwise ARI comparison (Figure \ref(fig:ari-fig),
Table \ref(tab:ari-table)) shows that K-means and RKM produce highly similar
partitions, with an average ARI of 0.587 and a maximum of 0.988. This is consistent
with their nearly identical silhouette scores and suggests that the two methods are
recovering the same underlying structure despite their different formulations. By
contrast, all method pairs involving PD or FPDC show very low average ARI values,
generally below 0.10, indicating that the probabilistic methods are recovering
fundamentally different partitions than K-means and RKM. The within-method ARI for
FPDC (average 0.271) also indicates moderate variability across FPDC solutions at
different $(k, q)$ settings, reflecting sensitivity to the tuning grid.

```{r sil-k-fig, fig.cap="Silhouette scores across candidate values of k for K-means and PD clustering, stratified by embedding dimension."}
plot_silhouette_vs_k(plot_df, methods = c("KMeans", "PD"))
```

Across all methods and embedding dimensions, the best solution by silhouette was
consistently found at $k = 2$, with the exception of FPDC on the 64-dimensional
embedding which selected $k = 3$ (Table \ref(tab:tbl-best-models)). The silhouette
profile for K-means across $k$ (Figure \ref(fig:sil-k-fig)) shows a sharp decline
after $k = 2$ in all embedding dimensions, confirming that the data support a broad
two-group partition most strongly. The flat, near-zero silhouette profile for PD across
the same range of $k$ further confirms that PD clustering did not find meaningful
structure at any cluster count.

## Stability Analysis

Bootstrap stability was assessed using a common design across all four clustering
methods. The 64, 128, and 256 dimensional user embeddings were each evaluated for
$k = 2, 3, 4$. For every combination of embedding and $k$, 15 bootstrap replicates
were generated by sampling 30% of users with replacement, and the clustering model was
re-fitted on each bootstrap subsample before being compared with the corresponding
full-data reference partition using ARI. Reduced K-means used reference models selected
over $q \in \{2, \ldots, 6\}$ and FPDC used reference models selected over
$q \in \{2, \ldots, 4\}$.

Figure \ref{fig:kmeans_bootstrap} shows the bootstrap stability profile for
K-means across the three embedding sizes. The 128 and 256 dimensional embeddings show
the same overall pattern: stability is highest at $k = 2$, remains moderate at $k = 3$,
and drops sharply at $k = 4$. The 64 dimensional embedding is less stable overall, with
only modest ARI at $k = 2$ and $k = 3$ and a similarly large decline at $k = 4$.
Taken together, these results suggest that K-means produces its most reproducible
partitions at lower cluster counts, especially for the higher dimensional embeddings,
while four-cluster solutions appear substantially less stable. This is consistent with
the internal validation results, where silhouette scores also peaked at $k = 2$.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/kmeans-stability.png}
    \caption{Bootstrap stability for K-means across embeddings and cluster counts.}
    \label{fig:kmeans_bootstrap}
\end{figure}

Figure \ref{fig:rkm_bootstrap} shows the bootstrap stability results for Reduced
K-means. The strongest stability appears for $k = 2$ in the 128 and 256 dimensional
embeddings, where mean ARI is high and clearly exceeds the values seen for the 64
dimensional embedding. For the 64 dimensional embedding, stability decreases noticeably
when moving from $k = 2$ to $k = 4$. Among the available results, RKM appears highly
stable at $k = 2$ and compares favorably with standard K-means, reinforcing the
conclusion from the silhouette analysis that both methods recover the same underlying
two-group structure.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/rkm-stability.png}
    \caption{Bootstrap stability for Reduced K-means across embeddings and cluster counts.}
    \label{fig:rkm_bootstrap}
\end{figure}


Figure \ref{fig:pd_bootstrap} presents the bootstrap stability results for PD
clustering. In contrast to K-means and RKM, ARI values are extremely close to zero
across nearly all settings, with only a small positive value for the 64-dimensional
embedding at $k = 2$. The 128 and 256 dimensional embeddings show near-zero or slightly
negative mean ARI throughout, indicating that bootstrap partitions do not reproduce the
full-data clustering in a consistent way. This result is consistent with the near-zero
silhouette scores reported in Table \ref{tab:tbl-best-models}, and together they
suggest that PD clustering failed to identify stable or well-separated structure in
these embeddings.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/pd-stability.png}
    \caption{Bootstrap stability for PD clustering across embeddings and cluster counts.}
    \label{fig:pd_bootstrap}
\end{figure}

Figure \ref{fig:fpdc_bootstrap} shows the bootstrap stability profile for FPDC.
As with PD clustering, mean ARI values are clustered very close to zero across all
embeddings and values of $k$, although estimates are slightly positive in most cases.
The 128 and 64 dimensional embeddings show a small local increase at $k = 3$, while the
256 dimensional embedding remains nearly flat across the tested cluster counts. This
pattern indicates that FPDC did not recover highly reproducible partitions under
bootstrap resampling, even though its solutions were marginally more stable than PD in
some settings. Combined with its poor silhouette scores and substantially higher
runtimes, FPDC was the least competitive method across all three evaluation dimensions
considered in this study.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/fpdc-stability.png}
    \caption{Bootstrap stability for FPDC across embeddings and cluster counts.}
    \label{fig:fpdc_bootstrap}
\end{figure}
