# Results

Structure:

1. Effect of embedding size: compare 64 vs. 128 vs. 256.
2. Algorithm comparison: k-means vs. PD vs. reduced k-means vs. FPDC.
3. Best model selection based on validation metrics.

[ADD LATER: ALL NUMERIC RESULTS.]

[ADD LATER: TABLES + PLOTS.]

[ADD LATER: OBSERVATIONS (E.G., "HIGHER DIMENSION IMPROVES SEPARATION BUT INCREASES VARIANCE").]


# `Report/sections/06-results.Rmd`

# Results

This section compares clustering performance across embedding dimensions and methods
using average silhouette width as the primary criterion, with the Calinski--Harabasz
(CH) index, runtime, and bootstrap stability as secondary criteria. All results are
generated directly from cached `.rds` outputs.

## Best model by method and embedding dimension

```{r tbl-best-models}
knitr::kable(
  best_table,
  booktabs = TRUE,
  caption = "Best clustering result for each method and embedding dimension."
)
```

Across all method--embedding combinations, the highest silhouette value in the cached results was obtained by `r as.character(overall_best$method)` using the `r overall_best$embedding_dim`-dimensional embeddings with `r overall_best$k` clusters`r if (!is.na(overall_best$q)) paste0(" and q = ", overall_best$q) else ""`. This configuration achieved a silhouette score of `r round(overall_best$silhouette, 3)` and a Calinski--Harabasz value of `r round(overall_best$ch, 2)`.


[HARD CODED OPTION FOR PARAGRPAH ABOVE]
Across all methods, the 128-dimensional embedding consistently produced the strongest
clustering structure as measured by silhouette width, followed by 256 and then 64
dimensions (Figure \@ref(fig:best-silhouette-fig), Table \@ref(tab:best-model-table)).
This pattern suggests that 64 dimensions may be insufficient to capture the full latent
structure of user preferences, while 256 dimensions may introduce enough noise or
redundancy to slightly reduce cluster separation. The 128-dimensional embedding appears
to offer the best balance between representational capacity and signal clarity for this
dataset.

## Effect of embedding size

```{r fig-best-silhouette, fig.cap="Best silhouette score attained by each clustering method within each embedding dimension."}
plot_best_silhouette_by_method_embedding(plot_df)
```

Within each embedding dimension, K-means and RKM achieved substantially higher
silhouette scores than PD and FPDC (Figure \@ref(fig:best-silhouette-fig),
Table \@ref(tab:best-model-table)). The best overall solution was K-means on the
128-dimensional embedding with $k = 2$, achieving a silhouette of 0.334 and CH of
139.51. RKM performed nearly identically to K-means across all settings, with silhouette
scores differing by no more than 0.003 at any embedding dimension, suggesting that the
additional dimensionality reduction step in RKM did not meaningfully alter the partition
structure recovered by K-means alone in this setting. Across all methods and embedding
dimensions, the best solution by silhouette was consistently found at $k = 2$, with the
exception of FPDC on the 64-dimensional embedding which selected $k = 3$. The silhouette
profile for K-means across $k$ (Figure \@ref(fig:sil-k-fig)) shows a sharp decline
after $k = 2$ in all embedding dimensions, confirming that the data support a broad
two-group partition most strongly.

## Algorithm comparison and computational cost

```{r fig-runtime-tradeoff, fig.cap="Runtime-performance tradeoff across all cached clustering models. Lower runtime and higher silhouette are preferred."}
plot_runtime_vs_silhouette(plot_df)
```

The runtime-performance tradeoff plot helps distinguish methods that are competitive in clustering quality from those that impose substantially higher computational cost. It should be interpreted jointly with the best-model table rather than as a standalone model-selection rule.

## Agreement across methods

```{r fig-ari-method-pairs, fig.cap="Average adjusted Rand index by method pair, computed within matched embedding dimensions and values of k."}
plot_ari_method_pairs(ari_df)
```

Adjusted Rand index (ARI) is used here as a stability and agreement measure across methods under comparable settings. Higher values indicate that two methods tend to recover more similar partitions when they are fit to the same embedding dimension and the same number of clusters.

## Tuning patterns for the baseline methods

```{r fig-silhouette-vs-k, fig.cap="Silhouette scores across candidate values of k for K-means and PD clustering, stratified by embedding dimension."}
plot_silhouette_vs_k(plot_df, methods = c("KMeans", "PD"))
```

This final plot provides a compact view of how the tuning parameter `k` changes the clustering quality for the two baseline methods across the embedding dimensions. If you need to shorten the section, this is the first figure I would move to the appendix.



