# Appendix {#sec-appendix}

The code used to reproduce all analyses, figures, and tables in this report is available at: \url{https://github.com/syedalijabir/math252-project}

## Additional tuning diagnostics

```{r fig-rkm-sensitivity, fig.cap="Reduced K-means silhouette values across candidate k and q settings.", include=FALSE}
plot_rkm_sensitivity(plot_df)
```


```{r tbl-ari-summary, include=FALSE}
## Agreement summary table
knitr::kable(
  format_ari_summary_table(ari_summary),
  booktabs = TRUE,
  caption = "Summary of pairwise adjusted Rand index values across method pairs."
)
```


<!-- \begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{figures/figure1_kmeans_macqueen_silhouette_vs_k.pdf}
    \caption{Average silhouette width across candidate cluster counts for MacQueen K-means. Silhouette peaks at 2 clusters in all three embedding dimensions, even though the elbow diagnostic selects 4 clusters.}
    \label{fig-kmeans-silhouette-k}
\end{figure} -->


\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.85\textwidth]{figures/appendix_figureA1_kmeans_wss_vs_k_cleaned.png}
    \caption{Within-cluster sum of squares for MacQueen K-means across candidate cluster counts. The elbow diagnostic selects 4 clusters in all three embedding dimensions.}
    \label{fig-kmeans-wss-appendix}
\end{figure}


\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.85\textwidth]{figures/appendix_figureA2_rkm_silhouette_vs_k_nf.png}
    \caption{Reduced K-means silhouette values across candidate cluster counts and retained-factor settings. The strongest region lies at 3 clusters if we look at the elbow technique, and 2 clusters if we use average silhouette width alone.}
    \label{fig-rkm-grid-appendix}
\end{figure}

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.85\textwidth]{figures/appendix_figureA3_probability_silhouette_pd_fpdc.png}
    \caption{Best probability silhouette across candidate cluster counts for PD clustering and FPDC. FPDC improves the soft-membership diagnostic relative to PD, but it still does not produce a strong hard partition.}
    \label{fig-prob-silhouette-appendix}
\end{figure}

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.85\textwidth]{figures/stability_all_methods_by_embedding.png}
    \caption{Bootstrap stability for all methods across embedding dimensions and cluster counts.}
    \label{fig-stability}
\end{figure}

<!-- \begin{figure}[htbp] -->
<!--     \centering -->
<!--     \includegraphics[width=0.9\textwidth]{figures/kmeans-stability.png} -->
<!--     \caption{Bootstrap stability for K-means across embedding dimensions and cluster counts.} -->
<!--     \label{fig-kmeans-stability} -->
<!-- \end{figure} -->

<!-- \begin{figure}[htbp] -->
<!--     \centering -->
<!--     \includegraphics[width=0.9\textwidth]{figures/rkm-stability.png} -->
<!--     \caption{Bootstrap stability for Reduced K-means across embedding dimensions and cluster counts.} -->
<!--     \label{fig-rkm-stability} -->
<!-- \end{figure} -->

<!-- \begin{figure}[htbp] -->
<!--     \centering -->
<!--     \includegraphics[width=0.9\textwidth]{figures/pd-stability.png} -->
<!--     \caption{Bootstrap stability for PD clustering across embedding dimensions and cluster counts.} -->
<!--     \label{fig-pd-stability} -->
<!-- \end{figure} -->

<!-- \begin{figure}[htbp] -->
<!--     \centering -->
<!--     \includegraphics[width=0.9\textwidth]{figures/fpdc-stability.png} -->
<!--     \caption{Bootstrap stability for FPDC across embedding dimensions and cluster counts.} -->
<!--     \label{fig-fpdc-stability} -->
<!-- \end{figure} -->

\begin{figure}
    \centering
    \includegraphics[width=0.45\textwidth]{figures/TuckerFactor.png}
    \caption{Tucker-factor diagnostic for the learned embeddings. The diagnostic increases smoothly as the number of retained factors grows and does not show a sharp elbow, which supports a compact retained-factor search range.}
    \label{fig-tucker-factor}
\end{figure}


\begin{table}
\centering
\small
\caption{FPDC sensitivity to the tuning parameter \(\nu\). Each row reports the best FPDC model within a fixed \(\nu\) value after excluding the inconsistent \(\nu=2\) pilot runs.}
\label{tbl-fpdc-nu-sensitivity}
\resizebox{\textwidth}{!}{
\begin{tabular}{rrrrrrrr}
\toprule
Embedding dimension & \(\nu\) & Clusters & Retained factors & Runtime (s) & Average Silhouette Width & Calinski--Harabasz index & Probability Silhouette \\
\midrule
64  & 6  & 2 & 4 & 241.74 & 0.012  & 70.55 & 0.143 \\
64  & 10 & 2 & 4 & 77.76  & 0.012  & 69.93 & 0.143 \\
128 & 6  & 4 & 2 & 188.16 & -0.115 & 56.43 & 0.136 \\
128 & 10 & 4 & 2 & 69.08  & -0.115 & 56.59 & 0.136 \\
256 & 6  & 3 & 2 & 170.20 & -0.056 & 63.41 & 0.146 \\
256 & 10 & 6 & 3 & 70.74  & -0.209 & 33.92 & 0.148 \\
\bottomrule
\end{tabular}
}
\end{table}




