# Appendix

You can see our entire code at our GitHub repository.

[WE SHOULD CITE THE PAPERS WE READ ABOUT THE DIFFERENT CLUSTERING METHODS!]

#`Report/sections/08-appendix.Rmd`



This appendix collects supporting diagnostics that informed model selection but were too detailed for the main results section.

## Additional tuning diagnostics

```{r fig-rkm-sensitivity, fig.cap="Reduced K-means silhouette values across candidate k and q settings."}
plot_rkm_sensitivity(plot_df)
```

```{r fig-prob-silhouette, fig.cap="Probability silhouette values for PD and FPDC across the candidate tuning grid."}
plot_prob_silhouette(plot_df)
```

## Agreement summary table

```{r tbl-ari-summary}
knitr::kable(
  format_ari_summary_table(ari_summary),
  booktabs = TRUE,
  caption = "Summary of pairwise adjusted Rand index values across method pairs."
)
```


\begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{figures/figure1_kmeans_macqueen_silhouette_vs_k.pdf}
    \caption{Average silhouette width across candidate cluster counts for MacQueen K-means. Silhouette peaks at 2 clusters in all three embedding dimensions, even though the elbow diagnostic selects 4 clusters.}
    \label{fig:kmeans-silhouette-k}
\end{figure}



\begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{figures/appendix_figureA1_kmeans_wss_vs_k_cleaned.pdf}
    \caption{Within-cluster sum of squares for MacQueen K-means across candidate cluster counts. The elbow diagnostic selects 4 clusters in all three embedding dimensions.}
    \label{fig:kmeans-wss-appendix}
\end{figure}


\begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{figures/appendix_figureA2_rkm_silhouette_vs_k_nf.pdf}
    \caption{Reduced K-means silhouette values across candidate cluster counts and retained-factor settings. The strongest region lies at 2 clusters.}
    \label{fig:rkm-grid-appendix}
\end{figure}

\begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{figures/appendix_figureA3_probability_silhouette_pd_fpdc.pdf}
    \caption{Best probability silhouette across candidate cluster counts for PD clustering and FPDC. FPDC improves the soft-membership diagnostic relative to PD, but it still does not produce a strong hard partition.}
    \label{fig:prob-silhouette-appendix}
\end{figure}

\begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{figures/kmeans-stability.png}
    \caption{Bootstrap stability for K-means across embedding dimensions and cluster counts.}
    \label{fig:kmeans-stability}
\end{figure}

\begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{figures/rkm-stability.png}
    \caption{Bootstrap stability for Reduced K-means across embedding dimensions and cluster counts.}
    \label{fig:rkm-stability}
\end{figure}

\begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{figures/pd-stability.png}
    \caption{Bootstrap stability for PD clustering across embedding dimensions and cluster counts.}
    \label{fig:pd-stability}
\end{figure}

\begin{figure}[htbp]
    \centering
    \includegraphics[width=\textwidth]{figures/fpdc-stability.png}
    \caption{Bootstrap stability for FPDC across embedding dimensions and cluster counts.}
    \label{fig:fpdc-stability}
\end{figure}

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.78\textwidth]{figures/TuckerFactor.png}
    \caption{Tucker-factor diagnostic for the learned embeddings. The diagnostic increases smoothly as the number of retained factors grows and does not show a sharp elbow, which supports a compact retained-factor search range.}
    \label{fig:tucker-factor}
\end{figure}


\begin{table}[htbp]
\centering
\small
\caption{FPDC sensitivity to the tuning parameter \(\nu\). Each row reports the best FPDC model within a fixed \(\nu\) value after excluding the inconsistent \(\nu=2\) pilot runs.}
\label{tab:fpdc-nu-sensitivity}
\resizebox{\textwidth}{!}{
\begin{tabular}{rrrrrrrr}
\toprule
Embedding dimension & \(\nu\) & Clusters & Retained factors & Runtime (s) & Average silhouette width & Calinski--Harabasz index & Probability silhouette \\
\midrule
64  & 6  & 2 & 4 & 241.74 & 0.012  & 70.55 & 0.143 \\
64  & 10 & 2 & 4 & 77.76  & 0.012  & 69.93 & 0.143 \\
128 & 6  & 4 & 2 & 188.16 & -0.115 & 56.43 & 0.136 \\
128 & 10 & 4 & 2 & 69.08  & -0.115 & 56.59 & 0.136 \\
256 & 6  & 3 & 2 & 170.20 & -0.056 & 63.41 & 0.146 \\
256 & 10 & 6 & 3 & 70.74  & -0.209 & 33.92 & 0.148 \\
\bottomrule
\end{tabular}
}
\end{table}



The full project code and supporting scripts are available in the GitHub repository. In the final version of the report, this appendix should also include the citations for the clustering methods and any supplementary implementation details that are too technical for the main text.
```



