# Appendix

You can see our entire code at our GitHub repository.

[WE SHOULD CITE THE PAPERS WE READ ABOUT THE DIFFERENT CLUSTERING METHODS!]

#`Report/sections/08-appendix.Rmd`



This appendix collects supporting diagnostics that informed model selection but were too detailed for the main results section.

## Additional tuning diagnostics

```{r fig-rkm-sensitivity, fig.cap="Reduced K-means silhouette values across candidate k and q settings."}
plot_rkm_sensitivity(plot_df)
```

```{r fig-prob-silhouette, fig.cap="Probability silhouette values for PD and FPDC across the candidate tuning grid."}
plot_prob_silhouette(plot_df)
```

## Agreement summary table

```{r tbl-ari-summary}
knitr::kable(
  format_ari_summary_table(ari_summary),
  booktabs = TRUE,
  caption = "Summary of pairwise adjusted Rand index values across method pairs."
)
```

The full project code and supporting scripts are available in the GitHub repository. In the final version of the report, this appendix should also include the citations for the clustering methods and any supplementary implementation details that are too technical for the main text.
```



